Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-10-29 paper/Response.tex

\documentclass[11pt]{article}
\usepackage[pdftex]{graphicx}
\usepackage{url}

\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9.25in}

\setlength{\voffset}{-0.0in}
\hoffset=-50pt
\topmargin=0pt
\headheight=0pt
\headsep=0pt
\evensidemargin=72pt % 1in+7pt

\setlength{\parskip}{0.5ex}
\setlength{\parindent}{1em}
\setlength{\floatsep}{1pt}
\setlength{\textfloatsep}{2pt}
\setlength{\intextsep}{2pt}

\begin{document}

{\it
October 29, 2010\\
Response to reviewers:  \\
We thank the reviewers for their careful reading of our paper.
Their suggestions definitely strengthen the paper and sparked new ideas
for improving our algorithms.
Below we list the Reviewers' comments verbatim.
Our responses are shown in italics. \\
Steve Plimpton and 
Karen Devine
}

\vspace{0.5 in}

Reviewer \#1: The submitted paper features a description and analysis of a
new MapReduce implementation based on MPI.  The features and limitations
of the implementation are covered with respect to other MapReduce
implementations.  The implementation is applied to several
graph-theoretic algorithms, and the performance of these applications
is compared with applications from an established library (Trilinos).

The introductory and supporting text are, overall, very good.  The
paper places MapReduce in its historical context and motivates its
usage.  The outline of strengths is helpful- important features
include support for out-of-core operations that other libraries (e.g.,
Trilinos) often cannot handle and the ability to use MPI directly in
addition to the MapReduce abstraction.  The paper then provides a good
overview of the composition of operations in a MapReduce application
and a summary of the out-of-core mechanisms available.  The figure
(Figure 1) used to provide a high-level schematic of MapReduce data
flows could be enhanced- for example, it should indicate node-local
and communication intensive operations.

All performance results are shown as scaling studies for a range of
problem sizes for the six graph algorithms.  The results show that the
MapReduce applications do not compete well with the reference
implementation (Trilinos), but the authors make clear the limitations
involved in using this high-level library which trades some
performance for usability.  Additionally, the implementation supports
very large data sets which can only be processed using its out-of-core
feature set.  The results are presented well.

Since improved performance for small and mid-range problems is not the
contribution of the paper, it could be improved by stressing other
aspects of the system.  Real-world out-of-core problems could be
surveyed.  Code size and complexity could be considered with usability
as a motivation.  Is the MapReduce-based graph algorithm library
easier to use than the Trilinos library?  Should higher-level user
applications be considered?  Can the MapReduce-based inefficiencies be
isolated more concisely?

The paper does not have a concise "Conclusion", only a "Lessons Learned"
section.

\vspace{0.5 in}

Reviewer \#2: This is a nice paper addressing an important current topic
with a good quality open-source implementation.  The overall presentation
is good: the description of algorithms and implementations is clear and the
performance evaluation is detailed. As the authors pointed out, the
performance of the MR-MPI implementations of some graph algorithms falls
behind the matrix-based counterparts so it would be interesting to further
investigate where the performance descrepency comes from and if it's
possible to optimize the algorithm and implementation to bridge the
performance gap.

{\it We instrumented our MR-MPI and Trilinos implementations of PageRank 
to show where time is spent during the PageRank iterations.  These results
are in Table 2.  We further instrumented the two most expensive steps of
the PageRank iterations, as shown in Tables 3 and 4.  These experiments
showed that much time is spent in reorganizing data from key-value pairs
to key-multivalue pairs.  In sections 4.1 and 5.2, we describe alternative
data layouts that can reduce the number of times the data must be 
reorganized to improve performance.}

\vspace{0.5 in}

Reviewer \#3: I appreciated the clear exposition of your map/reduce
framework, its decomposition into a number of independent and composable
stages, and the explanation of how out-of-core computations are handled.

I also liked that you presented the map/reduce implementation of a number
of different graph algorithms. However, I thought the presentation of these
algorithms could be improved in two ways:
- The text tends to just explaining what the code does, not why, nor why
  this solves the specific problem (4.2 was the worst culprit here - f I
understood correctly, the vertices get labeled by edge counts that are used
to give the edges a direction that helps identify triangles: if this is
correct a little diagram showing this would help...)
- The detailed algorithms (Figures 4-9) are a little too imprecise to
  follow at times. In particular, when multiple map/reduces are involved
the lack of names for the map/reduce operations and their inputs makes
things confusing. Please  be very explicit (after all the code itself is),
and Figure 4 at least actually had names for the map-reduce objects.

Your performance is quite low when compared to non map/reduce
implementations of the same algorithm. It would be interesting to know the
breakdown of execution time between map, shuffle and reduce phases to get a
better understanding of where your system spends its time (not necessarily
for every single performance measurement, but some representative or
overall measurements would be interesting).

{\it Reviewer \#2 made this same suggestion; please see the response above.}

Minor:
- 5.4: In Figure *13* (not 6)...

{\it Fixed; good catch.}
\end{document}

### 2010-10-30 paper/results.tex
Fixed one typo caught by reviewers. Started a response document Response.tex.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@649 3fc257bb-8550-0410-81b3-a5c103da83f7
Some RMAT-28 results.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@650 3fc257bb-8550-0410-81b3-a5c103da83f7

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline 
& \multicolumn{2}{|c|}{RMAT-20} & \multicolumn{2}{|c|}{RMAT-28} \\
& MR-MPI & Trilinos & MR-MPI & Trilinos  \\
\hline
Step 1 (Compute adjustment) & 0.0203 & 0.00030 &
&  \\
Step 2 (Multiply $A^T x$) & 0.1377 & 0.00455 &
&  \\
Step 3 (Apply adjustment) & 0.0015 & 0.00015 &
&  \\
Step 4 (Scale $y$)  & 0.0013 & 0.00014 &
& \\
Step 5 (Compute residual)  & 0.0207 & 0.00014 &
& \\
\hline
\end{tabular}
\caption{Detailed execution times (in seconds) per PageRank iteration
for each stage of the algorithm
in Figure~\ref{fig:pr2} using MR-MPI and Trilinos.}
\label{table:prdetail}
\end{center}
\end{table}

To understand where MR-MPI spends its effort during the PageRank computation,
we break down the operation times further for the most expensive 
step of the PageRank computation, Step 2; the results are in 
Table~\ref{table:prdetail2}.
In the table, we divide the collate operation into its
two parts:  an aggregate that involves parallel communication and a convert
that gathers local key-value pairs with matching keys into a key-multivalue
pair.  All interprocessor communication occurs in the aggregate operation.
While the time spent in aggregate is significant, it is small compared to
the time spent in the two convert operations.  Further experimentation
is needed to determine whether reorganizing the MapReduce data 
(as described in Section~\ref{subsec:graph_pagerank}) would reduce the
execution time of Step 2.  Organizing the data such that both the matrix
$A$ and vectors $x$ and $y$ are in the same MapReduce object would eliminate
the copy and add functions, as well as the first convert.  However, to update
the $y$ values at the completion of Step 2, either an additional convert
or an aggregate with greater amounts of data would be needed.

STILL NEED RMAT-28 RESULTS AND DISCUSSION OF I/O EFFECT.

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|r|r|}
\hline 
PageRank Step 2 & RMAT-20 & RMAT-28 \\
\hline
Copy $x$ to $y$ & 1.1\% &  \\
Add $A$ to $y$ & 3.0\% &  \\
Convert $y$ & 23.5\% &  \\
Reduce $y$ & 7.7\% & \\
Aggregate $y$ (first part of collate) & 24.9\% & \\
Convert $y$ (second part of collate) & 38.2\% & \\
Reduce $y$ & 1.6\% & \\
\hline
\end{tabular}
\caption{Percentage of execution time in each operation of Step 2 of the MR-MPI 
PageRank algorithm in Figure~\ref{fig:pr2}.}
\label{table:prdetail2}
\end{center}
\end{table}

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|r|r|}
\hline 
PageRank Step 5 & RMAT-20 & RMAT-28 \\
\hline
Add $y$ to $x$ & 2.2\% &  \\
Convert $x$ & 95.1\% &  \\
Reduce $x$ & 2.5\% & \\
MPI\_Allreduce & 0.2\% & \\
\hline
\end{tabular}
\caption{Percentage of execution time in each operation of Step 5 of the MR-MPI 
PageRank algorithm in  Figure~\ref{fig:pr2}.}
\label{table:prdetail5}
\end{center}
\end{table}
%This result is
%due to two factors: ({\it i}) a higher volume of communication in the
%MapReduce implementation (which comunicates one datum per graph edge
%in each iteration), compared to the Trilinos implementation which only
%communicates a volume of data proportional to the number of vertices,
%and ({\it ii}) out-of-core operations in MR-MPI were performed because
%of the page-size restriction, while all Trilinos operations were
%performed in-core.  
%

@@

In Figure~\ref{fig:cc}, we show the performance of the connected
component identification algorithm (Figure~\ref{fig:cc}).  A
comparison is made with a hybrid ``Giant Connected Component''
implementation using both Trilinos and MR-MPI.  Power law graphs often
have one or two very large components and many very small components.
The hybrid algorithm exploits this feature of the data by using
inexpensive breadth-first search (BFS) from the vertex with highest
degree to identify the largest components, followed by a more
expensive algorithm to identify the small components in the remainder
of the graph.  In our hybrid implementation, we intially perform
matrix-vector multiplications in Trilinos to perform a BFS, finding
components that include (in total) 70\% or more of the vertices.  We
then apply our MR-MPI algorithm to the remaining graph to identify the
small components.  This hybrid approach is quite effective, reducing
the execution time to identify all components of RMAT-20 and RMAT-24
by 80-99\%, as shown in Figure~\ref{f:cc}.  The benefit of MR-MPI is
seen, however, for RMAT-28, where the graph is too large to fit into
memory, and thus Trilinos cannot perform the initial BFS for our
hybrid algorithm.

>>>>

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline 
& \multicolumn{2}{|c|}{RMAT-20} & \multicolumn{2}{|c|}{RMAT-28} \\
& MR-MPI & Trilinos & MR-MPI & Trilinos  \\
\hline
Step 1 (Compute adjustment) & 0.0203 & 0.00030 & 17.53 & 0.0838 \\
Step 2 (Multiply $A^T x$) & 0.1377 & 0.00455 & 341.26 & 2.9239 \\
Step 3 (Apply adjustment) & 0.0015 & 0.00015 & 2.14 & 0.0494 \\
Step 4 (Scale $y$)  & 0.0013 & 0.00014 & 1.74 & 0.0359 \\
Step 5 (Compute residual)  & 0.0207 & 0.00014 & 21.20 & 0.0457 \\
\hline
\end{tabular}
\caption{Detailed execution times (in seconds) per PageRank iteration
for each stage of the algorithm
in Figure~\ref{fig:pr2} using MR-MPI and Trilinos.}
\label{table:prdetail}
\end{center}
\end{table}

To understand where MR-MPI spends its effort during the PageRank computation,
we break down the operation times further for the most expensive 
step of the PageRank computation, Step 2; the results are in 
Table~\ref{table:prdetail2}.
In the table, we divide the collate operation into its
two parts:  an aggregate that involves parallel communication and a convert
that gathers local key-value pairs with matching keys into a key-multivalue
pair.  All interprocessor communication occurs in the aggregate operation.
While the time spent in aggregate is significant, it is small compared to
the time spent in the two convert operations.  When out-of-core operations
are used for RMAT-28, the communication cost in the aggregate operation is
a smaller percentage of total time for Step 2, 
as other data manipulation operations such as convert and add 
take a relatively larger percentage of the total time.
Further experimentation
is needed to determine whether reorganizing the MapReduce data 
(as described in Section~\ref{subsec:graph_pagerank}) would reduce the
execution time of Step 2.  Organizing the data such that both the matrix
$A$ and vectors $x$ and $y$ are in the same MapReduce object would eliminate
the copy and add functions, as well as the first convert.  However, to update
the $y$ values at the completion of Step 2, either an additional convert
or an aggregate with greater amounts of data would be needed.

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|r|r|}
\hline 
PageRank Step 2 & RMAT-20 & RMAT-28 \\
\hline
Copy $x$ to $y$                       &  1.1\% &  0.4\%\\
Add $A$ to $y$                        &  3.0\% & 14.3\% \\
Convert $y$                           & 23.5\% & 43.6\% \\
Reduce $y$                            &  7.7\% &  5.8\% \\
Aggregate $y$ (first part of collate) & 24.9\% &  6.8\% \\
Convert $y$ (second part of collate)  & 38.2\% & 27.4\% \\
Reduce $y$                            &  1.6\% &  1.7\% \\
\hline
\end{tabular}
\caption{Percentage of execution time in each operation of Step 2 of the MR-MPI 
PageRank algorithm in Figure~\ref{fig:pr2}.}
\label{table:prdetail2}
\end{center}
\end{table}

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|r|r|}
\hline 
PageRank Step 5 & RMAT-20 & RMAT-28 \\
\hline
Add $y$ to $x$ &  2.2\% &  9.8\% \\
Convert $x$    & 95.1\% & 84.9\% \\
Reduce $x$     &  2.5\% &  5.3\% \\
MPI\_Allreduce &  0.2\% &  0.0\% \\
\hline
\end{tabular}
\caption{Percentage of execution time in each operation of Step 5 of the MR-MPI 
PageRank algorithm in  Figure~\ref{fig:pr2}.}
\label{table:prdetail5}
\end{center}
\end{table}
%This result is
%due to two factors: ({\it i}) a higher volume of communication in the
%MapReduce implementation (which comunicates one datum per graph edge
%in each iteration), compared to the Trilinos implementation which only
%communicates a volume of data proportional to the number of vertices,
%and ({\it ii}) out-of-core operations in MR-MPI were performed because
%of the page-size restriction, while all Trilinos operations were
%performed in-core.  
%

@@

In Figure~\ref{f:cc}, we show the performance of the connected
component identification algorithm (Figure~\ref{fig:cc}).  A
comparison is made with a hybrid ``Giant Connected Component''
implementation using both Trilinos and MR-MPI.  Power law graphs often
have one or two very large components and many very small components.
The hybrid algorithm exploits this feature of the data by using
inexpensive breadth-first search (BFS) from the vertex with highest
degree to identify the largest components, followed by a more
expensive algorithm to identify the small components in the remainder
of the graph.  In our hybrid implementation, we intially perform
matrix-vector multiplications in Trilinos to perform a BFS, finding
components that include (in total) 70\% or more of the vertices.  We
then apply our MR-MPI algorithm to the remaining graph to identify the
small components.  This hybrid approach is quite effective, reducing
the execution time to identify all components of RMAT-20 and RMAT-24
by 80-99\%, as shown in Figure~\ref{f:cc}.  The benefit of MR-MPI is
seen, however, for RMAT-28, where the graph is too large to fit into
memory, and thus Trilinos cannot perform the initial BFS for our
hybrid algorithm.
